### R/sl4hm: Simulated Likelihood for modelling household epidemics


[Yair Goldberg](https://yairgo.net.technion.ac.il/) and Rami Yaari

[R/sl4hm](https://github.com/yairgoldy/sl4hm) is an [R](https:/www.r-project.org) package. This package implements simulations of epidemics within households of children and adults and calculates the likelihood of a set of transmission parameters given a dataset of the
result of epidemics (number of infected children and adults) within households.

#### Installation

You can install it from its [GitHub repository](https://github.com/yairgoldy/sl4hm). You first need to install the [devtools](https://github.com/hadley/devtools) package.

```r
install.packages("devtools")
```

Then install R/sl4hm using the `install_github` function in the
[devtools](https://github.com/hadley/devtools) package.

```r
library(devtools)
install_github("yairgoldy/sl4hm")
```
### The data 

Loading the data used for the analysis and saving it to a csv file ca can be done as follows. 
```{r}
library(sl4hm)
household_data
write.csv(x = household_data,file = "household_data.csv")
```

The more detailed individualized-data level of the household data can be downloaded directly from this GitHub repository.

#### Example use

First create a dataset in which each row corresponds to a family and includes the number of adults and children and the index case.

```{r}

family_structure_data <- create_family_structure_data(n=1000)
head(family_structure_data,n = 5)

```

Then simulate the epidemic dynamic within each household, given
the transmission matrix for children and adults (beta) and a generation time distribution (P). The returned dataset includes the number of infected adults and children in each household.
```{r}
dat <- create_simulated_data(family_structure_data = family_structure_data,
                             beta_cc = 0.1,
                             beta_ca = 0.2,
                             beta_ac = 0.2,
                             beta_aa = 0.4,
                             P=sl4hm::load_P())
head(dat,n = 5)

                           
```
Calculate the likelihood surface for this dataset. While the simulated data can
be generated with any combination of beta values, the likelihood function assumes
a unique structure of beta in which beta_ac=beta_aa*delta, beta_ca=beta_aa*gamma,
and beta_cc=beta_aa*delta*gamma, where gamma and delta are non-negative coeficients
measuring the susceptibility and infectivity of children compared to adults.
The likelihood is calculated for a coarse grid. One can download a dictionary [from here](https://technionmail-my.sharepoint.com/:f:/g/personal/yairgo_technion_ac_il/EoMePBbpQkxLkju-mi51-ywB8pVeWgiEFY99IpHFtynjbw?e=dgVBUk) which enables calculating the likelihood on a finer grid. This dictionary holds the distribution of the number of infected for each family type and duration of epidemic. One can also create a dictionary using the function `family_dictionary`.
```{r}
lik <- calc_likelihood(dat)
head(lik,5)
require(ggplot2,quietly = T)

ggplot(lik,aes(x=gamma_val,y=delta_val,z=nll))+
  facet_wrap(~beta_aa,labeller="label_both", ncol=3)+
  geom_contour_filled(binwidth=5)+
  geom_contour(color = "black", size = 0.1,binwidth=5) +
  xlab("Child/Adult susceptablity")+
  ylab("Child/Adult infectivity")+
ggtitle("Likelihood Surface")+
 theme_bw()

```

Using the function `run_simulation` one can create multiple datasets and find the MLE for each dataset. This can be used for calculating bootstrap confidence intervals for the parameters.
